---
title: "covid19 qPCR analysis"
date: "23 April 2020"
output: html_document
---

```{r setup, include=FALSE}
options(stringsAsFactors = F)
knitr::opts_chunk$set(echo = TRUE, tidy = TRUE, warning = F, message = F, dpi = 200, fig.width=12, fig.height=8)
```

## organise libraries and files

```{r}
## packages
suppressPackageStartupMessages(require(tidyverse))
suppressPackageStartupMessages(require(janitor))
suppressPackageStartupMessages(require(chipPCR))

## working directory
workdir <- "C://Users/u4667515/Dropbox/Collab_Projects/Covid19"
setwd(workdir)

## arrange sample information

### RSB LC480
rsb_data_path <- file.path(workdir, "RawData_v2_standardised/")
rsb_files_rn <- dir(file.path(workdir, "RawData_v2_standardised/"), pattern = "raw-fluorescence")
rsb_files_ct <- dir(file.path(workdir, "RawData_v2_standardised/"), pattern = "Ct-values")
rsb_files_keyfile <- dir(file.path(workdir, "RawData_v2_standardised/"), pattern = "sample-sheet")

## JCSMR TaqMan
jcsmr_dir <- file.path(workdir, "RawData_JCSMR_v1/")
jcsmr_data_paths <- file.path(jcsmr_dir, dir(jcsmr_dir))
jcsmr_files_rn <- dir(jcsmr_data_paths, pattern = "Amplification-Data")
jcsmr_files_ct <- dir(jcsmr_data_paths, pattern = "Ct-values")
jcsmr_files_keyfile <- dir(jcsmr_data_paths, pattern = "sample-sheet")

### arrange sample information, fluorescence & ct values

keyfile_rsb <- tibble(filename = rsb_files_keyfile) %>% 
  mutate( file_contents = map(filename,~ read_delim(file.path(rsb_data_path, .), delim = '\t'))) %>%
  unnest %>% clean_names %>% 
  mutate(filename = paste(sapply(strsplit(filename, "_sample"), function(l) l[1]), well, sep='_')) %>% select(-well)

ct_rsb <- tibble(filename = rsb_files_ct) %>% 
  mutate( file_contents = map(filename,~ read_delim(file.path(rsb_data_path, .), delim = '\t'))) %>%
  unnest %>% clean_names %>%
  mutate(probe = ifelse(sapply(strsplit(filename, "_"), function(l) l[4]) == "SYBR", yes = "SYBR", no=probe )) %>% 
  mutate(filename = paste(sapply(strsplit(filename, "_Ct"), function(l) l[1]), well, sep='_')) %>%
  select(-well) %>%
  mutate(plate = sapply(strsplit(filename, "_"), function(l) l[3]))

data_rsb <- tibble(filename = rsb_files_rn) %>% 
  mutate( file_contents = map(filename,~ read_delim(file.path(rsb_data_path, .), delim = '\t'))) %>%
  unnest %>% clean_names %>%
  mutate(filename = paste(sapply(strsplit(filename, "_raw"), function(l) l[1]), well, sep='_')) %>%
  left_join(keyfile_rsb, by="filename") %>%
  na.omit %>%                         ## only clear empty wells
  left_join(ct_rsb, by="filename") %>%
  mutate(plate = sapply(strsplit(filename, "_"), function(l) l[3])) %>%
  filter( ct_value != 0 ) ## only wells with correct product

keyfile_jcsmr <- tibble(filename = jcsmr_files_keyfile) %>% 
  mutate( file_contents = map(filename,~ read_delim(file.path(jcsmr_data_paths, .), delim = ' ', col_names = F))) %>%
  unnest %>% clean_names %>% rename(well=x1, sample_name=x2, reaction_volume=x3, well_number=x4, dilution=x5) %>%
  mutate(filename = paste(sapply(strsplit(filename, "_sample"), function(l) l[1]), well_number, sep='_')) %>% 
  select(-well, -well_number)

ct_jcsmr <- tibble(filename = jcsmr_files_ct) %>% 
  mutate( file_contents = map(filename,~ read_csv(file.path(jcsmr_data_paths, .), skip = 36))) %>%
  unnest %>% clean_names %>%
  select(filename, well, sample_name, target_name, reporter, ct) %>%
  rename(probe=reporter) %>%
  mutate(type = sapply(strsplit(filename,"_Ct-values-"), function(l) l[2])) %>%
  mutate(filename = paste(sapply(strsplit(filename,"_Ct-values-"), function(l) l[1]), well, sep='_')) %>%
  mutate(type = sapply(strsplit(type, ".csv"), function(l) l[1])) %>%
  mutate(ct = as.numeric(paste(ifelse(ct == "Undetermined", yes = 0, no = ct)))) %>%
  spread(type, ct) %>%
  mutate(plate_name = paste(sapply(strsplit(filename, "_DS"), function(l) l[1]), "DS", sep='_'))

data_jcsmr <- tibble(filename = jcsmr_files_rn) %>% 
  mutate( file_contents = map(filename,~ read_csv(file.path(jcsmr_data_paths, .), skip = 36, 
            col_types = list(col_number(), col_number(), col_character(), col_number(), col_number())))) %>%
  unnest %>% clean_names %>% rename(fluorescence=rn) %>% select(-delta_rn) %>% na.omit %>%
  mutate(plate_name = sapply(strsplit(filename, "_Amp"), function(l) l[1])) %>%
  mutate(filename = paste(sapply(strsplit(filename, "_Amp"), function(l) l[1]), well, sep='_')) %>%
  left_join(keyfile_jcsmr, by="filename") %>%
  mutate(sample_name = ifelse(is.na(sample_name) == T, 
                              yes = ct_jcsmr$sample_name[match(filename, ct_jcsmr$filename)],
                              no = sample_name)) %>%
  mutate(sample_name = sapply(strsplit(sample_name, "_rep"), function(l) l[1]))
  

```

## Plot raw fluorescence

```{r, include=FALSE}
## use this section to plot individual plate fluorescence curves or machine-calculated Ct's to inspect
plate_ids <- unique(data_rsb$plate)[10]

# ggplot(subset(data_rsb, plate==plate_ids), aes(x=cycle, y=fluorescence, col=probe_amplicon, fill=probe_amplicon)) +
#   geom_point() + 
#   geom_vline(xintercept = c(33,38), linetype="dotted", size=0.5)+
#   facet_wrap(~sample_name, 'free_y') +
#   geom_smooth(formula = y ~ x^2)

# ggplot(data = subset(data_rsb, cycle ==2 & plate==plate_ids), 
#        aes(x=sample_name, y=ct_value, col=probe_amplicon, fill=probe_amplicon)) +
#   geom_point() + 
#   theme(axis.text.x = element_text(angle=45, hjust=1, size=8)) + 
#   scale_y_continuous(name = "Ct") +
#   scale_x_discrete(name = "Sample") + 
#   coord_cartesian(ylim = c(0,50)) +
#   geom_hline(yintercept = 38, linetype = "dashed")

plate_name = unique(data_jcsmr$plate_name)[1]

# ggplot(subset(data_jcsmr, plate_name==plate_name), aes(x=cycle, y=fluorescence, col=target_name, fill=target_name)) +
#   geom_point(size=1) + 
#   geom_vline(xintercept = c(33,38), linetype="dotted", size=0.5)+
#   facet_wrap(~sample_name, 'free_y') +
#   geom_smooth(formula = y ~ x^2)

# ggplot(data = subset(ct_jcsmr, plate_name==plate_name), 
#        aes(x=sample_name, y=default, col=target_name, fill=target_name)) +
#   geom_jitter(stat = 'identity') + 
#   theme(axis.text.x = element_text(angle=45, hjust=1, size=8)) + 
#   scale_y_continuous(name = "Ct (default)") +
#   scale_x_discrete(name = "Sample") + 
#   coord_cartesian(ylim = c(0,50)) +
#   geom_hline(yintercept = 38, linetype = "dashed")
# 
# ggplot(data = subset(ct_jcsmr, plate_name==plate_name), 
#        aes(x=sample_name, y=manual, col=target_name, fill=target_name)) +
#   geom_jitter(stat = 'identity') + 
#   theme(axis.text.x = element_text(angle=45, hjust=1, size=8)) + 
#   scale_y_continuous(name = "Ct (manual)") +
#   scale_x_discrete(name = "Sample") + 
#   coord_cartesian(ylim = c(0,50)) +
#   geom_hline(yintercept = 38, linetype = "dashed")

```

## chipPCR

#### pre-QC visualization

```{r}
## Organise fluorescence curves for target plate(s) e.g RSB plate 9
plate_ids <- unique(data_rsb$plate)[9]

a <- subset(data_rsb, plate == plate_ids)

my_list_rsb <- list()

for(i in unique(a$probe_amplicon)){
print(i)
  
  ## subset amplicons on plate to analyse
   b <- filter(a, probe_amplicon == i) %>%
    select(filename, cycle, fluorescence) %>%
    spread(filename, fluorescence) %>%
    as.data.frame
  
  ## give short form sample name
  sample_ids <- colnames(b)[2:ncol(b)]
  sample_ids <- paste(data_rsb$sample_name[match(sample_ids, data_rsb$filename)],i, sep='_')
  colnames(b)[2:ncol(b)] <- sample_ids
  
  ## QC plots
  # plot(MFIaggr(x=b)) ## plot Mean Fluorescence Intensity by cycle per target per plate
  # plotCurves(x=b[,1], y=b[,c(2:ncol(b))], type='l', CPP = TRUE) ## initial visual assessment with CPP QC
  
  ## keep all data in a list
  my_list_rsb[[i]] <- b

}

## e.g JCSMR
plate_name = unique(data_jcsmr$plate_name)[1]
a <- subset(data_jcsmr, plate_name==plate_name)

my_list_jcsmr <- list()

for(i in unique(a$target_name)){
print(i)
  b <- filter(a, target_name == i) %>%
  select(filename, cycle, fluorescence) %>%
  spread(filename, fluorescence) %>%
  as.data.frame
    
  ## give short form sample name
  sample_ids <- colnames(b)[2:ncol(b)]
  sample_ids <- paste(data_jcsmr$sample_name[match(sample_ids, data_jcsmr$filename)],i,sep='_')
  colnames(b)[2:ncol(b)] <- sample_ids
  
  ## QC plots
  # plot(MFIaggr(x=b)) ## plot Mean Fluorescence Intensity by cycle per target per plate
  # plotCurves(x=b[,1], y=b[,c(2:ncol(b))], type='l', CPP = TRUE) ## initial visual assessment with CPP QC
  
  my_list_jcsmr[[i]] <- b

}

```

#### RSB data analysis

Standardize data and calculate Ct using threshold method.

```{r}
par(mfrow=c(3,3))

CPP_list <- list()
th_list <- list()

for( i in names(my_list_rsb)){
  a <- my_list_rsb[[i]]
  
### CPP functions to pre-process data (e.g. smooth, normalize, background correction)
res.CPP <- apply(a[, -1], MARGIN = 2, function(l) {CPP(a[, 1], l, 
                                                       method="savgol",     ## Savitzky-Golay smoothing
                                                       trans=T,             ## background correction
                                                       method.reg="lmrob",  ## robust linear reg
                                                       bg.range = c(1,20),  ## set cycle range for background
                                                       bg.outliers = T,     ## remove outliers in background
                                                       method.norm = "none" ## no normalization
)[["y.norm"]]})

## calculate Cqs using threshold method = 0.1
th.cyc.CPP <- apply(res.CPP, 2, function(l) {th.cyc(a[, 1], l, r = 0.1)[1, 1]})

matplot(res.CPP, type = "l", xlab = "Cycle", ylab = "Fluorescence", main = paste("CPP\n",i))
abline(h = 0.1, lty = 2)

}
par(mfrow=c(1,1))
```

### Vazyme kit JCSMR data

The data from Vazyme kits are too noisy for reliable normalization (due to inability to diagnose aspecific ampification). Therefore, we cannot use the thresholding method in a consistent manner as above. Instead, we use the smoothed curves to interpolate first- and second-derivative maximums using the five-point stencil method.

We apply three filters to consider Ct values reliable:

1. Reactions with a max fluorescence exceeding a threshold correspoding to the smallest observation of sample quantiles generated from the distribution of fluorescence values across the plate (represented by vertical line on histogram). This threshold value is depicted as a horizontal red line on fluorescence curves.
2. FDM and SDM are not calculated during background range (<15 cycles)
3. |SDC - FDM| < 2 - in noisy reactions this difference is generally greater whereas successful reaction show a difference <0.5.

```{r}
par(mfrow=c(2,2))
cq <- NULL

for( i in names(my_list_jcsmr)){
  a <- my_list_jcsmr[[i]]
  
### CPP functions to pre-process data (e.g. smooth, normalize, remove background, remove outliers)
res.CPP <- apply(a[, -1], MARGIN = 2, function(l) {CPP(a[, 1], l,
                                                       method="supsmu",     ## Friedmann supersmoother
                                                       trans=T,             ## background slope correction
                                                       method.reg="lmrob",   ## robust linear reg
                                                       bg.range = c(1,22),   ## set cycle range for background
                                                       method.norm = "none", ## no normalization
                                                       bg.outliers = T      ### remove outliers in background
)[["y.norm"]]})

threshold = quantile(t(a[,-1]))[[1]]
hist(t(a[,-1]), breaks = 250, xlab = "Fluorescence", main = "Fluorescence distribution")
abline(v=threshold, col = "darkred")

## plot smoothed data
matplot(res.CPP, type = "l", xlab = "Cycle", ylab = "Fluorescence", main = paste("CPP\n",i))


  ## interpolate FDM and SDM from smoothed data and hold Cq values
  for( h in 1:ncol(res.CPP)){
    b <- clean_names(data.frame(cycle = a$cycle, res.CPP[, h]))
    res <- inder(b, smooth.method = NULL)
    summ <- summary(res, print = FALSE)
    out <- as.data.frame(t(summ))
    rownames(out) <- colnames(res.CPP)[h]
    out$max <- max(res[,2])
    out$threshold <- threshold
    
    test <- ifelse(out$max < out$threshold, yes = "bad", no=
                     ifelse( out$FDM < 15 | out$SDM < 15, yes = "bad" , no =
                               ifelse( abs(out$SDC-out$FDM)>2, yes = "bad", no = "Good!")))

    cq <- rbind(out, cq)
    
    col <- rainbow(4)
  
    ## plot all obtained metrics
    plot(res, 
  main=ifelse(test == "bad", yes = paste(colnames(res.CPP)[h], "removed!", sep='--'), no = colnames(res.CPP)[h]))
    abline(v=summ, col = col)
    text(summ["FDM"]-2, max(b$res_cpp_h)/1.2, paste0("FDM~", round(summ["FDM"], 2)), cex=0.8, col = col[1])
    text(summ["SDM"]-2, max(b$res_cpp_h)/2, paste0("SDM~", round(summ["SDM"], 2)), cex=0.8, col = col[2])
    text(summ["SDm"]-2, max(b$res_cpp_h)/3, paste0("SDm~", round(summ["SDm"], 2)), cex=0.8, col = col[3])
    text(summ["SDC"]-2, max(b$res_cpp_h)/4, paste0("SDC~", round(summ["SDC"], 2)), cex=0.8, col = col[4])
    abline(h=threshold, col = 'darkred')
  }


}

input <- as.data.frame(cq) %>% rownames_to_column(var = 'sample') %>% 
  mutate(ct = ifelse(FDM<=15 | SDM<=15, yes = 40, no = SDM)) %>%
  mutate(ct = ifelse(abs(SDC-FDM)<=2, yes = ct, no = 40)) %>%
  mutate(ct = ifelse(max < threshold, yes = 40, no = ct)) %>%
  mutate(sample = sapply(strsplit(sample, "\\."),function(l) l[1]))

lbls1 <- sapply(input$sample, function(l) { regexpr(l, pattern = "Internal Control") })
lbls1 <- lbls1[lbls1>1]
lbls2 <- sapply(input$sample, function(l) { regexpr(l, pattern = "N gene") })
lbls2 <- lbls2[lbls2>1]
lbls3 <- sapply(input$sample, function(l) { regexpr(l, pattern = "ORF1ab") })
lbls3 <- lbls3[lbls3>1]
lbls <- unlist(list(lbls1,lbls2,lbls3))

input <- input %>% 
  mutate(probe = substr(sample, start = lbls[match(sample, names(lbls))], stop = nchar(sample))) %>%
  mutate(sample = substr(sample, start = 1, stop= lbls[match(sample, names(lbls))] -2))

par(mfrow=c(1,1))

ggplot(data=input, aes(x=sample, y=ct, colour=sample)) +
  geom_jitter() + theme_bw() + 
  theme(axis.text.x = element_text(angle=45, hjust=1, size=7), legend.position = "none") +
  scale_y_continuous(name = "Ct") +
  scale_x_discrete(name = "Sample") +
  facet_wrap(~probe, ncol = 1)


```

